---
title: "Reproducible reports with Quarto"
toc: true
toc-location: right
---

# Introduction

## Current process

-   Typically people analyse, plot and write up in different programs.

-   Updates to data require manual updates to all the figures and tables
    as well as to text reporting results.

-   Graphs are saved to files and inserted into the final report.

-   Reordering figures, tables and equations means updating everywhere
    they are cross-referenced.

-   Submitting to a different journal requires reformatting the
    document.

-   Manual labour ..... error prone

## What is Quarto?

-   An open-source scientific and technical publishing system that
    allows you to create dynamic reproducible reports in a variety of
    output formats.

-   Next generation RMarkdown. Many new features, greater
    standardisation and more flexibility.

    -   Markdown is a lightweight markup language for creating formatted
        text using a plain-text editor. It was designed by [John Gruber
        in 2004](https://daringfireball.net/projects/markdown/)

    -   RMarkdown is an extension of Markdown that adds support for
        embedding R code chunks in Markdown documents. It was created by
        JJ Allaire and Yihui Xie early in 2014 but has been worked on by
        many others since.

-   Mutli-language: Python, R, Julia, and Observable.

-   Include auto-numbered equations, figures and tables that can be
    cross referenced

# Exercises

## Scenario

You are a data analyst for an international NGO. You have been asked to
produce a reproducible country profile report that analyses the
relationship between wealth (GDP per capita) and health (Life
Expectancy) over the last 50 years.

## The data

```{r}
#| echo: false
library(tidyverse)

gapminder <- read_tsv("data-raw/gapminder.tsv")



```

We will use data from [Gapminder](https://www.gapminder.org/). Gapminder
was founded in 2005 by Ola Rosling, Anna Rosling Rönnlund and Hans
Rosling. It aims to combat misguided global strategies rooted in
misconceptions by promoting a fact-based worldview grounded reliable
statistics. A subset of the data is
[gapminder.tsv](dat-raw/gapminder.tsv). It has `r nrow(gapminder)` rows
and `r ncol(gapminder)` columns (variables). The variables are:
`r paste(colnames(gapminder), collapse = ", ")`[^1]. There are
per-capita GDP and life expectancy data from
`r length(unique(gapminder$year))` years (from `r min(gapminder$year)`
to `r max(gapminder$year)`) for each of
`r length(unique(gapminder$country))` countries.

[^1]: Per-capita GDP (Gross domestic product) is given in units
    of [international
    dollars](https://en.wikipedia.org/wiki/Geary%E2%80%93Khamis_dollar),
    “a hypothetical unit of currency that has the same purchasing power
    parity that the U.S. dollar had in the United States at a given
    point in time” – 2005, in this case.

```{r}
#| echo: false
knitr::kable(gapminder) |> 
  kableExtra::kable_styling() |> 
  kableExtra::scroll_box(height = "200px")
```

## Create a new Quarto document

### Create a new project

🎬 File \| New Project \| New Directory \| Quarto Project

-   Browse to an appropriate place and give your project a name.

-   I used `ngo-report`

-   Choose Engine: Knitr

-   Turn visual markdown editor off (for now)

### Change some RStudio settings

🎬 Tools \| Global Options

-   General:

    -   Turn off the three "Restore ...." options

    -   Turn "Save workspace to .RData on exit" to Never

-   R Markdown:

    -   Turn "Show output preview in:" to Viewer pane

### Your project contains

-   `ngo-report.RProj` - what makes the folder an RStudio project

-   `_quarto.yml` - the configuration file

-   `ngo-report.md` - the main document containing some template text

The code is normal functioning code. You can run it interactively as you
write your report.

🎬 Place your cursor in the code chunk and run it (Ctrl-Enter or use the
green run button)

🎬 Hit Render (Ctrl-Shift-K)

Note that `ngo-report.html` is created and opened in the Viewer pane

## Edit the YAML header

🎬 Add your name, and a title. Also add the engine and format

``` yml
---
title: "The relationship between Gross Domestic Product per capita and Life Expectancy."
author: "Emma Rand"
engine: knitr
format:
  html
---
```

### Add some content

🎬 Add a code chunk (Ctrl-Shift-I) for a simple graph.

```{r}
#| echo: fenced
hist(rnorm(100))
```

🎬 Hit Render (Ctrl-Shift-K)

### Edit the YAML header: default code chunk options

🎬 Set some default code chunk options. I recommend these for reports

``` yml
---
title: "The relationship between Gross Domestic Product per capita and Life Expectancy."
author: "Emma Rand"
engine: knitr
format:
  html
execute:
  echo: false
  include: true
  error: false
  message: false
  warning: false

---
```

🚨 Indentation matters in YAML. Make sure the `execute` options are
indented under `execute`.

🎬 Hit Render (Ctrl-Shift-K)

Code chunk options

-   `echo: false` code will not be included in output
-   `include: true` output will be included
-   `error: false` halt render if a code error occurs
-   `message: false` messages and warnings will not be included
-   `warning: false`

🎬 Experiment with change the options and re-rendering. Try misspelling
`rnorm` and switching between `error: true` and `error: false` to see
the error option in action.

## Adding content

🎬 Delete everything except the YAML header.

🎬 Add headings for: Introduction, Methods, Results, Discussion, and
References

🎬 Add a code chunk for package loading

```{r}
#| echo: fenced
#| label: load-packages
library(tidyverse)
```

`#| label: load-packages` is a code chunk label. These are useful (here)
and essential (later) for cross-referencing.

### An Introduction

🎬 Add this test to the Introduction

> There is a relationship between wealth and health. We will explore
> this relationship using the relationship between Gross Domestic
> Product (GDP) per capita and Life Expectancy across countries and over
> time.

🎬 Make the "per capita" italic

🎬 Make "Gross Domestic Product (GDP)" a link to
<https://en.wikipedia.org/wiki/Gross_domestic_product>

### Data import

🎬 Make a folder called `data-raw`

🎬 Save [gapminder.tsv](data-raw/gapminder.tsv) to `data-raw`

🎬 Add a code chunk with:

```{r}
#| echo: fenced
#| label: import-data
file <- "data-raw/gapminder.tsv"
gapminder <- read_tsv(file)
```

🎬 Run chunks interactively

🎬 Render

### Methods text

🎬 Add this text to Methods:

> We took per capita GDP and life expectancy data from 12 years (from
> 1952 to 2007) for each of 142 countries. Analysis was carried out with
> R and tidyverse packages.

It would be good not to have "hard coded" those numbers in the text.
What if we get more data. Or we just misremember or mis-type the
numbers?

## Inline code: reproducible reporting

-   Inline code is how you include a variable value, like a sample size,
    mean or statistical result, in a section of text.

-   In fact, any code output can be inserted directly into the text of a
    .qmd file using inline code.

Inline code goes between `` `r `` and `` ` ``.

For example by writing:

`` `r sqrt(2)` ``

you will get:

`r sqrt(2)` \### Extract summary information for use in inline code

🎬 Add a code chunk with:

```{r}
#| echo: fenced
#| label: extract-summary
# extract values for inline reporting
n_years <- length(unique(gapminder$year))
min_year <- min(gapminder$year)
max_year <- max(gapminder$year)
n_counties <- length(unique(gapminder$country))

```

🎬 Replace numbers in the Methods text with inline code:

> We took per capita GDP and life expectancy data from `` `r n_years` ``
> years (from `` `r min_year` `` to `` `r max_year` ``) for each of
> `` `r n_counties` `` countries.

🎬 Render

Teach: Headers (# Introduction), bolding (**key findings**), and lists
(Study goals).

Teach: The YAML header (Add a Title: "Global Health Trends", Author, and
toc: true).

Part 2: Integrating Data (0:30 – 1:00) Goal: Show how code and text live
together.

The Task: Load the data and pull out specific stats for the text.

Scenario Action:

Insert a code chunk to load library(tidyverse) and library(gapminder).

Filter the data to the most recent year (2007).

Teach: Inline Code. "The average life expectancy across all countries in
2007 was r mean(gapminder_2007\$lifeExp) years." (This is a huge "Aha!"
moment for admin staff who manually update reports).

```{r}
#| label: data-prep
#| fence: true

# Filter the dataset to the most recent year for our report
gapminder_2007 <- gapminder |> 
  filter(year == 2007)
```

Part 3: Tables and Visuals (1:00 – 1:30) Goal: Demonstrate professional
output.

The Task: Create a summary table and a chart.

Scenario Action:

The Table: Use knitr::kable() or the gt package to show the "Top 5
Countries by Life Expectancy."

Teach: Chunk options like #\| echo: false (hiding the code so the
manager only sees the table).

The Plot: Create a ggplot scatterplot of GDP vs. Life Expectancy,
colored by Continent.

Teach: Figure captions and cross-referencing ("As seen in @fig-gdp...").

Part 4: The "Power User" Finish (1:30 – 2:00) Goal: Show why Quarto is
better than Word.

The Task: Citations and Formats.

Scenario Action:

Citations: Add a references.bib file. Cite Hans Rosling or the Gapminder
foundation within the text.

Rendering: Render the report to HTML. Then, with one click, render it to
Word (for the boss who tracks changes) or PDF (for the formal
publication).

Bonus (if time permits): Parameters. Change the country parameter in the
YAML to "Canada" or "Kenya" and watch the whole report update
automatically for that specific country.

Pages made with R [@R-core], Quarto [@Allaire_Quarto_2026], `knitr`
[@knitr1; @knitr2; @knitr3], `kableExtra` [@kableExtra]

# References
